R Notebook

### 2019-03-11 analysis/INLA/test-INLA-sensor.Rmd
messy trial code with INLA... but some of it works now!

```{r}
testdat <- tidy_survey_sets(raw, c("SYN HS","SYN QCS","SYN WCVI","SYN WCHG"), years = c(2015, 2016, 2017, 2018))
#testdat <- gfplot:::interp_survey_bathymetry(testdat)
glimpse(testdat)
View(testdat[[1]])
```

```{r}
# get_ssids()
sensor_syn <- get_sensor_data_trawl(ssid = c(1, 3, 4, 16), spread_attributes = TRUE)
# View(sensor_syn)
saveRDS(sensor_syn, "data-cache/allsynoptic_sensor_07mar19_time2.rds")
# write.csv(sensor, "synoptic_3_sensor_data_07mar19.csv")
# sensor <- read_csv("data-cache/synoptic_3_sensor_data_07mar19.csv", col_types = cols(do_mlpL_avg = col_double(), 
#                     do_mlpL_max = col_double(), do_mlpL_min = col_double(), salinity_PSU_avg = col_double(), salinity_PSU_max = col_double(), 
#                     salinity_PSU_min = col_double()), na = "NA")
sensor <- readRDS("data-cache/allsynoptic_sensor_07mar19_time2.rds")
View(sensor)
```

```{r}
dat <- left_join(testdat[[1]], sensor, by="fishing_event_id", suffix = c("", ".sensor"))
dat$year_f <- factor(dat$year)

View(dat)

dat <- dat %>% mutate(do_range = do_mlpL_max - do_mlpL_min, salinity_range = salinity_PSU_max -
  salinity_PSU_min, temp_range = temperature_C_max - temperature_C_min, depth_range = depth_m_max - depth_m_min)
glimpse(dat)
```

@@

plot(-log(depth) ~ do_mlpL_avg,  cex = depth_range/35, data = dat) 
plot(-log(depth) ~ temperature_C_avg,  cex = depth_range/35, data = dat)
plot(-log(depth) ~ salinity_PSU_max,  cex = depth_range/35, data = dat)

@@

ggplot(dat2, aes(x=depth, y=temperature_C_avg)) + 
  geom_point(alpha=0.25) + 
  geom_point(data=subset(dat2, temp_range>1), alpha=0.5) + 
  geom_errorbar(data=subset(dat2, temp_range>1),aes(ymin=temperature_C_min, ymax=temperature_C_max),alpha= 0.75, colour="blue") + 
  geom_errorbarh(data=subset(dat2, temp_range>1),aes(xmin=depth_m_min, xmax=depth_m_max),alpha= 0.75, colour="red") + 
  ylim(2,12) + 
  facet_wrap(~survey_desc, scales = "free")
```

@@

```{r}
scaled <- scale_survey_predictors(dat)
# glimpse(scaled)
synhs <- scaled %>% filter(do_mlpL_avg!="NA") %>% filter (density!="NA") 
glimpse(synhs)
```

```{r}
ggplot(synhs, aes(depth)) + geom_histogram() 
ggplot(synhs, aes(density*1000)) + geom_histogram() 
ggplot(synhs, aes(depth, (density*1000))) + geom_point() 
#ggplot(synhs, aes(depth_m, log(density*1000+1))) + geom_point() 
```

@@

@@

####
names(d) <- tolower(names(d))
d$species_common_name <- tolower(d$species_common_name)
d$species_science_name <- tolower(d$species_science_name)
d$year <- lubridate::year(d$trip_start_date)

sp <- filter(d, species_common_name %in% "pacific ocean perch") %>% 
  filter(!is.na(catch_weight)) %>% 
  filter(year %in% seq(2004, 2012, 2))

dat <- sp
dat <- filter(dat, start_lon > -128, start_lon < 125, start_lat > 48, start_lat < 50.3)
nrow(dat)

dat <- filter(dat, fe_bottom_water_temperature > 1, fe_bottom_water_temperature < 12,
  !is.na(fe_bottom_water_temp_depth), !is.na(fe_bottom_water_temperature))
nrow(dat)

dat <- select(dat, year, start_lon, start_lat, catch_weight, fe_bottom_water_temp_depth) %>% 
  rename(X = start_lon, Y = start_lat)

# Make a design matrix where the first year is the intercept
# ...
dm <- matrix(nrow = nrow(dat), ncol = 2)
dm[,1] <- 1
dm[,2] <- dat$fe_bottom_water_temp_depth

synhs$year_f2 <- synhs$year_f
synhs$year_f3 <- synhs$year_f

f1 <- temperature_C2 ~ depth_scaled + depth_scaled2 + 
                 f(year_f, model = "rw2") +
                 f(year_f2, depth_m, model = "rw2") +
                 f(year_f3, depth2, model = "rw2")

I1 <- inla(f1, 
           control.predictor = list(compute = TRUE),
           family = "gamma",
           data = synhs
           )
summary(I1)

@@

# Let us plot the 2 trends.
# Here is how we access them:
Yearsm <- I1$summary.random$Year
Daysm  <- I1$summary.random$DayInYear

# The rest is simple plotting
#par(mfrow = c(1,2), mar = c(5,5,2,2), cex.lab = 1.5)
plot(Yearsm[,1:2], type='l',
     xlab = 'Year', 
     ylab = 'Smoother',
     ylim = c(-0.2, 0.2) )
abline(h=0, lty=3)
lines(Yearsm[, c(1, 4)], lty=2)
lines(Yearsm[, c(1, 6)], lty=2)
text(1988, 0.2, "A", cex = 1.5)

plot(Daysm[,1:2], type='l',
     xlab = 'DayInYear', 
     ylab = 'Smoother',
     ylim = c(-0.5, 0.6) )
abline(h=0, lty=3)
lines(Daysm[, c(1, 4)], lty=2)
lines(Daysm[, c(1, 6)], lty=2)
text(0, 0.6, "B", cex = 1.5)

####################################

@@

###################
# TRY WITH interaction
inla.list.models()
# te = two-dimensional smoother

library(dplyr)

PB$Year2 <- PB$Year
PB$DayInYear2 <- PB$DayInYear

PB2 <- PB %>% filter(Year > 1991)

U <- 1.5

hyper.prec = list(theta = list(
  prior = "pc.prec", 
  param = c(U, 0.05)))

f3 <- Movement ~ fRepro + 
  f(Year, 
    model = "rw2",
    hyper = hyper.prec) +
  f(DayInYear, Year2,
    model = "rw2",
    hyper = hyper.prec) +  
  f(BearN, model = "iid")

I3 <- inla(f3, 
           control.predictor = list(compute = TRUE),
           family = "gamma",
           data = PB2
)
summary(I3)
Fit3 <- I3$summary.fitted.values[,"mean"]

# Access the trends
Yearsm <- I3$summary.random$Year
Daysm  <- I3$summary.random$DayInYear

# And plot the trends
#par(mfrow = c(1,2), mar = c(5,5,2,2), cex.lab = 1.5)
Time <- Yearsm[,1] #* sd(PB$Year) + mean(PB$Year)
plot(x = Time,
     y = Yearsm[,2], type='l',
     xlab = 'Year', 
     ylab = 'Trend',
     ylim = c(-0.5, 0.5) )
abline(h=0, lty=3)
lines(x = Time, y = Yearsm[, c(4)], lty=2)
lines(x = Time, y = Yearsm[, c(6)], lty=2)
text(1988, 0.5, "A", cex = 1.5)

@@

>>>>

```{r}
testdat <- tidy_survey_sets(raw, c("SYN HS","SYN QCS","SYN WCVI","SYN WCHG"), years = c(2015, 2016, 2017, 2018))
testdat <- gfplot:::interp_survey_bathymetry(testdat)
glimpse(testdat)
#View(testdat[[1]])
```

```{r}
# get_ssids()
# sensor_syn <- get_sensor_data_trawl(ssid = c(1, 3, 4, 16), spread_attributes = TRUE)
# View(sensor_syn)
# saveRDS(sensor_syn, "data-cache/allsynoptic_sensor_07mar19_time2.rds")
# write.csv(sensor, "synoptic_3_sensor_data_07mar19.csv")
# sensor <- read_csv("data-cache/synoptic_3_sensor_data_07mar19.csv", col_types = cols(do_mlpL_avg = col_double(), 
#                     do_mlpL_max = col_double(), do_mlpL_min = col_double(), salinity_PSU_avg = col_double(), salinity_PSU_max = col_double(), 
#                     salinity_PSU_min = col_double()), na = "NA")
sensor <- readRDS("data-cache/allsynoptic_sensor_07mar19_time2.rds")
# View(sensor)
```

```{r}
dat <- left_join(testdat[[1]], sensor, by="fishing_event_id", suffix = c("", ".sensor"))
# dat <- left_join(testdat, sensor, by="fishing_event_id", suffix = c("", ".sensor"))
dat$year_f <- factor(dat$year)
# View(dat)
dat <- dat %>% mutate(do_range = do_mlpL_max - do_mlpL_min, salinity_range = salinity_PSU_max -
  salinity_PSU_min, temp_range = temperature_C_max - temperature_C_min, depth_range = depth_m_max - depth_m_min)
# glimpse(dat)
```

@@

plot(-log(depth) ~ do_mlpL_avg,  cex = depth_range/35, data = dat) 
plot(-log(depth) ~ temperature_C_avg,  cex = depth_range/35, data = dat)
plot(-log(depth) ~ salinity_PSU_max,  cex = depth_range/35, data = dat)
```

@@

ggplot(dat2, aes(x=depth, y=temperature_C_avg)) + 
  geom_point(alpha=0.25) + 
  geom_point(data=subset(dat2, temp_range>1), alpha=0.5) + 
  geom_errorbar(data=subset(dat2, temp_range>1),aes(ymin=temperature_C_min, ymax=temperature_C_max),alpha= 0.75, colour="blue") + 
  geom_errorbarh(data=subset(dat2, temp_range>1),aes(xmin=depth_m_min, xmax=depth_m_max),alpha= 0.75, colour="red") + 
  ylim(2,12) + 
  facet_grid(rows = vars(year), cols = vars(ssid), scales = "free")
```

@@

```{r}
scaled <- scale_survey_predictors(dat)
# glimpse(scaled)
temp_dat <- scaled %>% filter (density!="NA") %>% filter ( ssid =="1") %>% filter(temperature_C_avg!="NA") # %>% filter(year==2017)
# glimpse(temp_dat)
# do_dat <- scaled %>% filter (density!="NA") %>% filter ( ssid =="1") %>%filter(do_mlpL_avg!="NA")
#pg <- readRDS("prediction-grids/QCSpred-grid-interp-cell-width-2.rds")
#pg <- as.array(pg)

```{r}
ggplot(sens_dat, aes(depth)) + geom_histogram() 
ggplot(sens_dat, aes(density*1000)) + geom_histogram() 
ggplot(sens_dat, aes(depth, (density*1000))) + geom_point() 
#ggplot(synhs, aes(depth_m, log(density*1000+1))) + geom_point() 
```

@@

```{r}
fit_inla_sensor <- function(dat, response = "temperature_C_avg",
                     n_knots = 50,
                     family = "binomial",
                     max_edge = c(20, 100),
                     kmeans = FALSE,
                     plot = FALSE,
                     fit_model = TRUE,
                     extend = list(n = 8, offset = -0.1),
                     offset = c(5, 25),
                     cutoff = 10,
                     include_depth = TRUE,
                     verbose = FALSE,
                     debug = FALSE,
                     trials = 10) {
  d <- dat
  coords <- as.matrix(unique(d[, c("X", "Y")]))

  # use kmeans() to calculate centers:
  if (kmeans) {
    km <- stats::kmeans(x = coords, centers = n_knots)
    mesh <- INLA::inla.mesh.create(km$centers,
      extend = extend
    )
  } else {
    bnd <- INLA::inla.nonconvex.hull(coords)
    mesh <- INLA::inla.mesh.2d(
      offset = offset,
      boundary = bnd,
      max.edge = max_edge,
      cutoff = cutoff
    )
  }

  message("INLA max_edge = ", "c(", max_edge[[1]], ", ", max_edge[[2]], ")")
  if (plot) {
    graphics::plot(mesh, asp = 1)
    graphics::points(coords, col = "red")
  }

  spde <- INLA::inla.spde2.matern(mesh, alpha = 3 / 2)
  dat <- data.frame(
    y = d[, response, drop = TRUE],
    xcoo = d$X,
    ycoo = d$Y
  )

  dat$intercept <- 1

  if (include_depth) {
    dat$depth <- d$depth_scaled
    dat$depth2 <- d$depth_scaled2
  }

  # construct index for ar1 model:
  iset <- INLA::inla.spde.make.index(
    "i2D",
    n.spde = mesh$n
  )

  # make the covariates:
  X.1 <- dat[, -c(1:3), drop = FALSE]
  covar_names <- colnames(X.1)
  XX.list <- as.list(X.1)
  effect.list <- list()
  effect.list[[1]] <- c(iset)
  for (Z in seq_len(ncol(X.1)))
    effect.list[[Z + 1]] <- XX.list[[Z]]
  names(effect.list) <- c("1", covar_names)

  A <- INLA::inla.spde.make.A(
    mesh = mesh,
    loc = cbind(dat$xcoo, dat$ycoo)
  )
  A.list <- list()
  A.list[[1]] <- A
  for (Z in seq_len(ncol(X.1)))
    A.list[[Z + 1]] <- 1

  # make projection points stack:
  sdat <- INLA::inla.stack(
    tag = "stdata",
    data = list(y = dat$y),
    A = A.list,
    effects = effect.list
  )

  formula <- as.formula(paste(
    "y ~ -1 +",
    paste(covar_names, collapse = "+"),
    "+ f(i2D, model=spde)"
  ))

  if (fit_model) {
    try_fitting <- TRUE
    trial_i <- 1
    while (try_fitting && trial_i <= trials) {
      model <- try(INLA::inla(formula,
        family = family,
        data = INLA::inla.stack.data(sdat),
        control.predictor = list(
          compute = TRUE,
          A = INLA::inla.stack.A(sdat)
        ),
        control.fixed = list(
          mean = 0, prec = 1 / (5^2),
          mean.intercept = 0, prec.intercept = 1 / (20^2)
        ),
        control.compute = list(config = TRUE),
        verbose = verbose,
        debug = debug,
        keep = FALSE
      ))
      trial_i <- trial_i + 1
      if (class(model) != "try-error") {
        try_fitting <- FALSE
      } else {
        # if (trial_i > 5) browser()
        message("INLA error. Trying again.")
      }
    }
    if (class(model) == "try-error") {
      model <- NA
    }
  } else {
    model <- NA
  }
  list(
    model = model, mesh = mesh, spde = spde, data = dat,
    formula = formula, iset = iset
  )
}
```

```{r}
obj <- fit_inla_sensor(temp_dat, response = "temperature_C_avg",
                       n_knots = 50,
                       family = "gaussian",
                       max_edge = c(20, 100),
                       kmeans = FALSE,
                       plot = FALSE,
                       fit_model = TRUE,
                       extend = list(n = 8, offset = -0.1),
                       offset = c(5, 25),
                       cutoff = 10,
                       include_depth = TRUE,
                       verbose = FALSE,
                       debug = FALSE,
                       trials = 2)

pg <- gfplot:::make_prediction_grid(filter(temp_dat, year %in% c(2016, 2017)), survey ="SYN QCS")$grid

cell_area <- gfplot:::make_prediction_grid(filter(temp_dat, year %in% c(2016, 2017)), survey = "SYN QCS")$cell_area

pred_temp <- predict_inla_sens(obj=obj, pred_grid = pg , samples = 100L,
                         include_depth = TRUE)
```

Project onto grid:

```{r posterior-draws}
pg_one_year <- filter(pg, year == min(pg$year))
predicted <- predict_inla_sens(obj = obj, pred_grid = pg_one_year, samples = 100, include_depth = TRUE)
```

@@

# alternative to function, hard code

samples = 10L
pred_grid <- pg

predict_inla_sens <- function(obj, pred_grid, samples = 100L,
                         include_depth = TRUE) {

  mesh <- obj$mesh
  model <- obj$model
  iset <- obj$iset

  inla.mcmc <- INLA::inla.posterior.sample(n = samples, model)

  # get indices of various effects:
  latent_names <- rownames(inla.mcmc[[1]]$latent)
  re_indx <- grep("^i2D", latent_names)
  if (include_depth) {
    depth_fe_indx1 <- grep("^depth$", latent_names)
    depth_fe_indx2 <- grep("^depth2$", latent_names)
  }
  intercept_indx <- grep("^intercept$", latent_names)

  # read in locations and knots, from projection matrix
  grid_locs <- pred_grid[, c("X", "Y")]

  # projections will be stored as an array:
  projected_latent_grid <-
    array(0, dim = c(nrow(grid_locs), samples))
  proj_matrix <- INLA::inla.spde.make.A(mesh, loc = as.matrix(grid_locs))

  # do MCMC projections:
  indx <- which(iset$i2D.group == 1)
  for (i in seq_len(samples)) {

    if (include_depth) {
      depth_effects <-
        pred_grid[, "depth_scaled", drop = TRUE] * inla.mcmc[[i]]$latent[depth_fe_indx1] +
        pred_grid[, "depth_scaled2", drop = TRUE] * inla.mcmc[[i]]$latent[depth_fe_indx2]
    } else {
      depth_effects <- 0
    }

    projected_latent_grid[, i] <-
      as.numeric(proj_matrix %*% inla.mcmc[[i]]$latent[re_indx][indx]) +
      inla.mcmc[[i]]$latent[intercept_indx] +
      depth_effects
  }

  projected_latent_grid
}

Calculate probability * Gamma median values on the grid:

```{r prediction-grid-calcs}
pg_one_year <- filter(pg, year == min(pg$year))
attributes(predicted)
# out <- stats::plogis(p_bin) * exp(p_pos)
out_med <- apply(predicted, c(1), median)
# out_bin <- stats::plogis(p_bin)
# out_bin <- apply(out_bin, c(1, 3), median)
dim(predicted)

pg_all <- bind_rows(replicate(1, pg_one_year, simplify = FALSE))

out_long <- reshape2::melt(out_med)
pg_all$pred <- out_long$value
```

Spatiotemporal plot:

```{r spatio-temporal-plot, fig.width=9, fig.height=9}
coast <- gfplot:::load_coastline(range(d$lon), range(d$lat), utm_zone = 9)
isobath <- gfplot:::load_isobath(range(d$lon), range(d$lat),
  bath = c(100, 200, 500), utm_zone = 9
)
map_padding <- c(-5, 5)

make_plot <- function(dat, fill = "sqrt(pred)") {
  g <- ggplot(dat, aes(X * 10, Y * 10)) +
    geom_raster(aes_string(fill = fill)) +
    scale_fill_viridis_c(option = "C") +
    #facet_wrap(~ year) +
    geom_polygon(
      data = coast, aes_string(x = "X", y = "Y", group = "PID"),
      fill = "grey80"
    ) +
    geom_point(
      data = filter(d, present == 1),
      aes(x = X * 10, y = Y * 10, size = density), inherit.aes = FALSE,
      pch = 21, col = "white", alpha = 0.3
    ) +
    geom_point(
      data = filter(d, present == 0),
      aes(x = X * 10, y = Y * 10), inherit.aes = FALSE,
      pch = 4, col = "white", alpha = 0.2
    )

  g <- g + geom_path(
    data = isobath, aes_string(
      x = "X", y = "Y",
      group = "paste(PID, SID)"
    ),
    inherit.aes = FALSE, lwd = 0.4, col = "grey70", alpha = 0.4
  )

  g <- g + xlab("UTM 9N Easting (km)") + ylab("UTM 9N Northing (km)") +
    scale_size(range = c(0, 9)) +
    guides(size = FALSE, fill = FALSE) +
    theme_pbs() +
    coord_equal(
      expand = FALSE, xlim = range(pg_all$X * 10) + map_padding,
      ylim = range(pg_all$Y * 10) + map_padding
    )

  g
}
# make_plot(pg_all)
median(pg_all$pred)

make_plot(pg_all, "pred") + 
  scale_fill_gradient2(
    midpoint = median(pg_all$pred),
    low = scales::muted("blue"), 
    mid = "white",
    high = scales::muted("red"))

make_plot(pg_all, "depth_scaled") + 
  scale_fill_gradient2(
    midpoint = 0,
    high = scales::muted("blue"), 
    mid = "white",
    low = scales::muted("red"))
```

Run using Zuur's polar bear code as template for letting relationship between temp and depth change through time...

```{r}
temp_dat$year_f2 <- temp_dat$year_f
temp_dat$year_f3 <- temp_dat$year_f

f1 <- temperature_C_avg ~ depth_scaled + depth_scaled2 + 
                 f(year_f, model = "rw2") +
                 f(year_f2, depth_scaled, model = "rw2") +
                 f(year_f3, depth_scaled2, model = "rw2")

I1 <- inla(f1, 
           control.predictor = list(compute = TRUE),
           family = "gaussian",
           data = temp_dat
           )
summary(I1)

@@

# Let us plot the 2 trends.
# Here is how we access them:
Yearsm <- I1$summary.random$year_f
Depthre  <- I1$summary.random$year_f2
Depth2re  <- I1$summary.random$year_f3
```

```{r}
# The rest is simple plotting
#par(mfrow = c(1,2), mar = c(5,5,2,2), cex.lab = 1.5)
plot(Yearsm[,1:2], type='l',
     xlab = 'Year', 
     ylab = 'Smoother',
     ylim = c(-0.2, 0.2) )
abline(h=0, lty=3)
lines(Yearsm[, c(1, 4)], lty=2)
lines(Yearsm[, c(1, 6)], lty=2)
text(1988, 0.2, "A", cex = 1.5)

plot(Depthre[,1:2], type='l',
     xlab = 'Depth', 
     ylab = 'Smoother',
     ylim = c(-0.5, 0.6) )
abline(h=0, lty=3)
lines(Depthre[, c(1, 4)], lty=2)
lines(Depthre[, c(1, 6)], lty=2)
text(0, 0.6, "B", cex = 1.5)
```

```{r}
####################################

@@

@@

Sean's code that I dont' remember where i found...

####
# names(d) <- tolower(names(d))
# d$species_common_name <- tolower(d$species_common_name)
# d$species_science_name <- tolower(d$species_science_name)
# d$year <- lubridate::year(d$trip_start_date)
# 
# sp <- filter(d, species_common_name %in% "pacific ocean perch") %>% 
#   filter(!is.na(catch_weight)) %>% 
#   filter(year %in% seq(2004, 2012, 2))
# 
# dat <- sp
# dat <- filter(dat, start_lon > -128, start_lon < 125, start_lat > 48, start_lat < 50.3)
# nrow(dat)
# 
# dat <- filter(dat, fe_bottom_water_temperature > 1, fe_bottom_water_temperature < 12,
#   !is.na(fe_bottom_water_temp_depth), !is.na(fe_bottom_water_temperature))
# nrow(dat)
# 
# dat <- select(dat, year, start_lon, start_lat, catch_weight, fe_bottom_water_temp_depth) %>% 
#   rename(X = start_lon, Y = start_lat)

dat <- temp_dat

# Make a design matrix where the first year is the intercept
# ...
dm <- matrix(nrow = nrow(dat), ncol = 2)
dm[,1] <- 1

#########

dm[,2] <- dat$fe_bottom_water_temp_depth

########
